---
title: "Custom helpers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom helpers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

While the chores package provides a number of pre-engineered helpers for R package development, one can use helpers for all sorts of coding tasks in R, from interactive data analysis to authoring with Quarto, or even for coding tasks in languages other than R! All you need to set up your own helper is a markdown file. This vignette provides guidance on how to write your own helpers to help you with repetitive, hard-to-automate tasks.

```{r setup}
library(chores)
```

## What are helpers for?

When building a custom chore helper, you should first ask yourself a few questions to help decide whether a chore helper is the right tool for the job.

**Can this be easily automated without an LLM?** LLMs are quite good at edge-case rich tasks with "squishy," hard-to-evaluate output. Compared to many other pieces of software, LLMs are incredibly flexible and able to handle edge cases without extensive engineering. At the same time, their output is not deterministic and always requires verification. For your given problem, would you be better off spending the time you'd use to write and revise a prompt and then verifying every output it makes from then on instead just writing some R code and unit tests?

Users interface with helpers via the active selection. **Would a selection provide enough context for the helper to do its job?** That is, besides the fixed system prompt attached to your helper, the only information the LLM has about your problem is the text that you've selected with your cursor. Would the model need access to a whole file, or a whole project, in order to do its job? In that case, some interface other than a helper may be a better fit.

Chore helpers return output by replacing, prefixing, or suffixing the current selection. **Is this the right place for output in your use case?** If output ultimately needs to be cut and pasted into a different file, for example, helpers may not be the right tool.

Finally, **is this kind of output immediately verifiable?** The first element to consider here is how much stuff the model ultimately generates. A few sentences or a couple lines of code can be quickly audited and confirmed to be sound. The second element to consider is whether the output can be programmatically checked. For example, does generated code run without error (or, at least, is it syntactically valid)?

Answering these questions can help you better understand whether your problem is best solved with a helper or, instead, by a human or "normal" code or some other LLM interface.

## The prompt directory

The easiest way to write a new chore helper that's loaded every time you load R is to add a markdown file to the _prompt directory_. The prompt directory is a folder of markdown files that serves as a library of possible helpers. By default, the prompt directory lives at `~/.config/chores`, but that default can be changed by adding a `.chores_dir` option in your `.Rprofile` using `options(.chores_dir = some/dir/`). `directory_path()` returns the path to the directory, `directory_set()` changes it, `directory_list()` enumerates all of the prompts that currently live in it, and `directory_load()` registers/refreshes all of the prompts in the directory with the chores package.

To create a new helper, add a prompt to chores' prompt directory with `prompt_new()`. You'll need to supply a `chore` (a single keyword describing what the helper does, like `"roxygen"`) and an `interface` describing how the helper will interact with the selection (one of `"replace"`, `"prefix"`, or `"suffix"`). You can also "pre-fill" the contents of the prompt by supplying a file path with the `contents` argument.

For example, running:

```{r}
#| eval: false
prompt_new("proofread", "replace")
prompt_new("summarize", "prefix")
```

Would result in a prompt directory that looks like:

```
/
|-- .config/
|   |-- chores/
|       |-- proofread-replace.md
|       |-- summarize-prefix.md
```

In that case, chores would register two custom helpers when you call `library(chores)` (or `directory_load()`. One of them is for the `"proofread"` chore and will replace the selected text with a proofread version (according to the instructions contained in the markdown file itself). The other is for the chore `"summarize"` and will prefix the selected text with a summarized version (again, according to the markdown file's instructions).

* Files without a `.md` extension are ignored.
* Files with a `.md` extension must contain only one hyphen in their filename, and the text following the hyphen must be one of `replace`, `prefix`, or `suffix`.

The best way to register helpers for your own personal use is via the prompt directory. However, if you intend to share helpers with others, you may be interested in creating a helper extension package. (If you do so, please add them to the chores gallery! Check it out at `vignette("gallery", package = "chores")`.)

## Extension packages

chores extension packages allow you to more flexibly share helper prompts with others. Putting together a helper extension package is straightforward:

* Place one markdown file per new chore helper in `inst/prompts/`. This folder will take the same format as the prompt directory described above.
* Place a call to `chores::directory_load()` in the package's `.onLoad()`, referencing the extension package's `system.file("prompts", package = "yourpackage")`. This will automatically register your package's prompts with chores when the extension is loaded.

For an example chores extension, see [simonpcouch/chorespable](https://github.com/simonpcouch/chorespable).

chores extension packages also allow you to document what your helpers are for and how they tend to behave in context; situate your documentation files at `?chore_helper`, replacing `chore` with your new chore helper's name. Then, with your package loaded, users can view a high-level description of the helper's behavior and a gallery of examples. See `?cli_helper` for an example helper help-page, with source code [here](https://github.com/simonpcouch/chores/blob/main/R/doc-chores-cli.R).

## Using others' custom helpers

chores is designed to make it as easy as possible to always have the helpers you need on hand.

**To use others' custom helpers that aren't situated in extension packages**, use `prompt_new()` with a `contents` argument. For example, Hannah Frick wrote a helper to transition R code chunks from R Markdown-style chunk headers to Quarto's yaml syntax and uploaded it as a GitHub Gist [here](https://gist.github.com/hfrick/1ca8fc2cb2a4409b743e8120c6cc2223#file-quartochunk-replace-md). To use her helper, we could write:

```r
prompt_new(
  "quartochunk", 
  "replace", 
  contents = "https://gist.githubusercontent.com/hfrick/1ca8fc2cb2a4409b743e8120c6cc2223/raw/a9703edfbd4e83839af0278c33add1b33e243d02/quartochunk-replace.md"
)
```

After running that code, the `quartochunk` helper will be available to you every time you trigger the chores addin.

**To use others' custom helpers from extension packages**, simply load the chores extension in your `.Rprofile`. You might use `usethis::edit_r_profile()` to open the file, then drop in the following line:

```r
library(choresextensionname)
```

Then, restart R, and that package's helpers will always be available to you when you trigger the chores addin.
